### 2021-05-20 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
Further edits to ch1 and 2. Added images for examples and spy plots of jacobi mats.

\bstodoinline{Summarise the ECMWF approach, and how this work is then a stepping stone to understanding more about multi dim OPs on complex domains.}

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. It is useful to be able to transform between these cartesian coordinates and the spherical coordinates ($\varphi, \theta$). Throughout this document we will use the convention that the spherical coordinate angles be defined by
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi.
}
where $\rho(z) := \sqrt{1-z^2}$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}:
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}. \label{eqn:clmdef}
} 
The spherical harmonics as defined are then orthonormal with respect to the inner product
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in $\R^3$.

@@

Using \bsrefeqns{eqn:shrecx}{eqn:shrecz}, we have that the Jacobi matrices take the following block-tridiagonal form (i.e. have block-bandwidths $(1,1)$):
\bseqn{
	J_x &= 
		\begin{pmatrix}
			B_{x,0} & A_{x,0} & & & & \\
			C_{x,1} & B_{x,1} & A_{x,1} & & & \\
			& C_{x,2} & B_{x,2} & A_{x,2}  & & & \\
			& & C_{x,3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}, \\
	J_y &= 
		\begin{pmatrix}
			B_{y,0} & A_{y,0} & & & & \\
			C_{y,1} & B_{y,1} & A_{y,1} & & & \\
			& C_{y,2} & B_{y,2} & A_{y,2}  & & & \\
			& & C_{y,3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}, \\
	J_z &= 
		\begin{pmatrix}
			B_{z,0} & A_{z,0} & & & & \\
			C_{z,1} & B_{z,1} & A_{z,1} & & & \\
			& C_{z,2} & B_{z,2} & A_{z,2}  & & & \\
			& & C_{z,3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}.
}
$J_x$ has subblock-bandwidths $(2,2)$, where the blocks for $l \in \No$ are given by:
\bseqn{
	A_{x,l} &:= 
		\begin{pmatrix}
			\alpha_{l,-l,3} & 0 & \alpha_{l,-l,4} & & \\
			& \ddots & \ddots & \ddots & \\
			& & \alpha_{l,l,3} & 0 & \alpha_{l,l,4} \\
		\end{pmatrix} \in \R^{(2l+1)\times(2l+3)}, \\
	B_{x,l} &:= 0 \in \R^{(2l+1)\times(2l+1)} \\
	C_{x,l} &:= 
		\begin{pmatrix}
			\alpha_{l,-l,2} & & \\
			0 & \ddots & \\
			\alpha_{l,-l+2,1} & \ddots & \alpha_{l,l-2,2} \\
			& \ddots & 0 \\
			& & \alpha_{l,l,1}
		\end{pmatrix} \in \R^{(2l+1)\times(2l-1)} \quad (l \ne 0).
}
$J_y$ also has subblock-bandwidths $(2,2)$, where the blocks for $l \in \No$ are given by:
\bseqn{
	A_{y,l} &:= 
		\begin{pmatrix}
			\beta_{l,-l,3} & 0 & \beta_{l,-l,4} & & \\
			& \ddots & \ddots & \ddots & \\
			& & \beta_{l,l,3} & 0 & \beta_{l,l,4} \\
		\end{pmatrix} \in \C^{(2l+1)\times(2l+3)}, \\
	B_{y,l} &:= 0 \in \R^{(2l+1)\times(2l+1)} \\
	C_{y,l} &:= 
		\begin{pmatrix}
			\beta_{l,-l,2} & & \\
			0 & \ddots & \\
			\beta_{l,-l+2,1} & \ddots & \beta_{l,l-2,2} \\
			& \ddots & 0 \\
			& & \beta_{l,l,1}
		\end{pmatrix} \in \C^{(2l+1)\times(2l-1)} \quad (l \ne 0).
}
Finally, $J_z$ has subblock-bandwidths $(1,1)$, where the blocks for $l \in \No$ are given by:
\bseqn{
	A_{z,l} &:= 
		\begin{pmatrix}
			0 & \gamma_{l,-l,2} & 0 & & \\
			& \ddots & \ddots & \ddots & \\
			& & 0 &  \gamma_{l,l,2} & 0 \\
		\end{pmatrix} \in \R^{(2l+1)\times(2l+3)}, \\
	B_{z,l} &:= 0 \in \R^{(2l+1)\times(2l+1)} \\
	C_{z,l} &:= 
		\begin{pmatrix}
			0 & & \\
			\gamma_{l,-l+1,1} & \ddots & \\
			0 & \ddots & 0 \\
			& \ddots & \gamma_{l,l-1,1} \\
			& & 0
		\end{pmatrix} \in \R^{(2l+1)\times(2l-1)} \quad (l \ne 0).
}

@@

	A_{l,m,5} &:= {1 \over l(l+1)}

@@

			(-1)^m \: i \: d_{l,\abs{m}} \: {\clm \over 2 c_l^{m-1}} &\text{if } m \le 0

@@

	A_{l,m,6} &:= {1 \over l(l+1)}

@@

			- i \: d_{l,m} \: {\clm \over 2 c_l^{m+1}} &\text{if } m \ge 0 \\
			(-1)^{m+1} \: i \: d_{l,\abs{m} - 1} \: {c_l^{m+1} \over 2 \clm} &\text{if } m < 0

@@

	B_{l,m,j} &:= i (-1)^{j+1} A_{l,m,j} \quad \text{for } j = 1,\dots,6, \\

@@

	\Gamma_{l,m,3} &:= {(-1)^m \: i \: m \over l(l+1)}

@@

	Fix $l \in \N$ and $m \in \Z$ s.t. $-l \le m \le l$. The proof we will use is to directly calculate the nonzero coefficients in each expansion using inner products. To this end, note that
\bseqn{
	x \gradYlm = \sum_{l'=1}^{l+1} \sum_{m'=-l'}^{l'} \calA_{l',m'} \gradYlmprime + \calA^\perp_{l',m'} \gradpYlmprime
}
where the coefficients are give by
\bseqn{
	\calA_{l',m'} = {\ip<x \gradYlm, \gradYlmprime> \over \norm{\gradYlmprime}^2} = {1 \over l'(l'+1)} \ip<x \gradYlm, \gradYlmprime> \\
	\calA^\perp_{l',m'} = {\ip<x \gradYlm, \gradpYlmprime> \over \norm{\gradpYlmprime}^2} = {1 \over l'(l'+1)} \ip<x \gradYlm, \gradpYlmprime>.
}
The aim will be to show that the only nonzero coefficients $\calA_{l',m'}$, $\calA^\perp_{l',m'}$ match the $A_{l,m,j}$ for $j=1,\dots,6$ in the Lemma. Now, using a change of variable, we have that
\bseqnnumber{
	&\ip<x \gradYlm, \gradYlmprime> \nonumber \\
	&= \iint_\Omega \costheta \sinphi \: \gradYlm \cdot {\gradYlmprime}^* \sinphi \: \D\phi \: \D\theta \nonumber \\
	&= \clm \clmprime \Big( \int_0^{2pi} \eimtheta \eimthetaprime \costheta \: \D\theta \Big) \nonumber \\
	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ \big[ \abs{m} z \jaclmabs{l}{m}(z) - \rho(z)^2 \jaclmabsdz{l}{m}(z) \big] \big[ \abs{m'} z \jaclmabs{l'}{m'}(z) - \rho(z)^2 \jaclmabs{l'}{m'}(z) \big] \nonumber \\
	&\quadeight + mm' \jaclmabs{l}{m}(z) \jaclmabs{l'}{m'}(z) \Big\} \: \rho(z)^{\abs{m}+\abs{m'}-1} \: \D z \Big) \nonumber \\
	&= \clm \clmprime \pi (\delta_{m',m-1} + \delta_{m',m+1}) \nonumber \\
	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ \jaclmabs{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m}+\abs{m'}-1} \big(mm' + \abs{m}\abs{m'} z^2 \big) \nonumber \\
	&\quadeight - \abs{m} z \jaclmabs{l}{m} \jaclmabsdz{l'}{m'} \rho^{\abs{m}+\abs{m'}+1} \nonumber \\
	&\quadeight - \abs{m'} z \jaclmabsdz{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m}+\abs{m'}+1} \nonumber  \\
	&\quadeight + \jaclmabsdz{l}{m} \jaclmabsdz{l'}{m'} \rho^{\abs{m}+\abs{m'}+3} \Big\} \: \D z \Big). \label{eqn:lemmaVSHrecurrencesproof}
}
We can conclude that the only possible nonzero coefficients are when $m' = m \pm 1$, and so it is left to evaluate the integral that is left at \bsrefeqn{eqn:lemmaVSHrecurrencesproof} when $m'$ takes either value. Since we need to take account of the fact that $m$ can be negative, to simplify our argument we make use of the fact that $m' = m - 1$ for $m < 0$ is analogous to the case $m' = m+1$ for $m > 0$, and vice versa.

@@

	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ 2 i \, m \, m' \, z \jaclmabs{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m} + \abs{m'} - 1} \nonumber \\

@@

	&\quadeight - i m' \jaclmabs{l}{m} \jaclmabsdz{l'}{m'} \rho^{\abs{m} + \abs{m'} +1} \Big\} \: \D z \Big). \label{eqn:lemmaVSHrecurrencesproof4}

@@

Once again, we can conclude that $\calA^\perp_{l',m'}$ are zero for $m' \notin \{m-1, m+1\}$. It is left to evaluate the integral that is left at \bsrefeqn{eqn:lemmaVSHrecurrencesproof4} when $m'$ takes either $m\pm 1$ value. First, consider the case that $m > 0$ and $m' = m - 1$. The integral is then, using integration by parts:

@@

	&\int_{-1}^{1} \Big\{ i \, \big[ m \abs{m-1} + \abs{m} (m-1) \big] \, z \jaclmabs{l}{m} \jaclmabs{l'}{m-1} \rho^{\abs{m} + \abs{m-1} - 1} \nonumber \\
	&\quadfour - i m \jaclmabs{l}{m} \jaclmabsdz{l'}{m-1} \rho^{\abs{m} + \abs{m-1} + 1} \nonumber \\
	&\quadfour - i (m-1) \jaclmabsdz{l}{m} \jaclmabs{l'}{m-1} \rho^{m + \abs{m-1} + 1} \Big\} \: \D z \nonumber \\

@@

	&= - \int_{-1}^{1} i \; d_{l',\abs{m}-1} \; \jaclm{l}{m} \jaclm{l'}{m} \rho^{2m} \: \D z \nonumber \\

@@

	&\int_{-1}^{1} \Big\{ i \, \big[ m \abs{m+1} + \abs{m} (m+1) \big] \, z \jaclmabs{l}{m} \jaclmabs{l'}{m+1} \rho^{\abs{m} + \abs{m+1} - 1} \nonumber \\
	&\quadfour - i m \jaclmabs{l}{m} \jaclmabsdz{l'}{m+1} \rho^{\abs{m} + \abs{m+1} + 1} \nonumber \\
	&\quadfour - i (m+1) \jaclmabsdz{l}{m} \jaclmabs{l'}{m+1} \rho^{\abs{m} + \abs{m+1} + 1} \Big\} \: \D z \nonumber \\

@@

	&= - \int_{-1}^{1} i \jaclmabs{l}{m} \jaclmbracketdz{l'}{\abs{m} - 1} \rho^{2\abs{m}} \: \D z \nonumber \\
	&= - \int_{-1}^{1} i \; d_{l',\abs{m}-1} \; \jaclmabs{l}{m} \jaclmabs{l'}{m} \rho^{2\abs{m}} \: \D z \nonumber \\
	&= - i \; {d_{l,\abs{m}-1} \over 2 \pi (c_l^{\abs{m}})^2} \; \delta_{l',l}. \label{eqn:lemmaVSHrecurrencesproof6}

@@

where the coefficients are give by

@@

The integral over $\theta$ in \bsrefeqn{eqn:lemmaVSHrecurrencesproofz1} is simply $2 \; \pi \; \delta_{m',m}$ showing the above is zero when $m' \ne m$, and thus so will the coefficients $\calC_{l',m'}$ be when this is the case. Focusing on the second integral (over $z$) when $m' = m$, this becomes:

@@

We can see here in \bsrefeqn{eqn:lemmaVSHrecurrencesproofz3} that the nonzero coefficients $\calC_{l',m'}$ are those described in the Lemma as $\Gamma_{l,m,1}$ and $\Gamma_{l,m,3}$.

Next, we also have that
\bseqnnumber{
	&\ip<z \gradYlm, \gradpYlmprime> \nonumber \\
	&= \iint_\Omega \cosphi \: \gradYlm \cdot {\gradpYlmprime}^* \sinphi \: \D\varphi \: \D\theta \nonumber \\
	&= \clm \clmprime \Big( \int_0^{2pi} \eimtheta \eimthetaprime \: \D\theta \Big) \nonumber \\
	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ i m' z \jaclmabs{l'}{m'}(z) \rho(z)^{\abs{m} + \abs{m'} - 2} \big[ \abs{m} z \jaclmabs{l}{m}(z) - \rho(z)^2 \jaclmabsdz{l}{m}(z) \big] \nonumber \\
	&\quadeight + i m z \jaclmabs{l}{m}(z) \rho(z)^{\abs{m} + \abs{m'} - 2} \big[ \abs{m'} z \jaclmabs{l'}{m'}(z) - \rho(z)^2 \jaclmabs{l'}{m'}(z) \big] \Big\} \: \D z \Big) \nonumber \\
	&= \clm c_{l'}^{m} 2 \pi \delta_{m',m} \: \int_{-1}^{1} \Big\{ 2i m\abs{m} z^2 \jaclmabs{l}{m}(z) \jaclmabs{l'}{m}(z) \rho(z)^{2\abs{m} - 2} \nonumber \\
	&\quadeight \quadtwo - i m z \jaclmabs{l}{m}(z) \jaclmabsdz{l'}{m}(z) \rho(z)^{2\abs{m}} \nonumber \\
	&\quadeight \quadtwo - i m z \jaclmabsdz{l}{m}(z) \jaclmabs{l'}{m}(z) \rho(z)^{2\abs{m}} \Big\} \: \D z \nonumber \\
	&= \clm c_{l'}^{m} 2 \pi \delta_{m',m} \: \int_{-1}^{1} i m z^2 \jaclmabs{l}{m}(z) \jaclmabs{l'}{m}(z) \rho(z)^{2\abs{m}} \: \D z \nonumber \\
	&= \delta_{l',l} \delta_{m',m} i m {(\clm)^2 \over (\clmprime)^2} \nonumber \\
	&= \delta_{l',l} \delta_{m',m} i m (-1)^m. \label{eqn:lemmaVSHrecurrencesproofz4}
}
Again, we can see here in \bsrefeqn{eqn:lemmaVSHrecurrencesproofz4} that the coefficients $\calC^\perp_{l',m'}$ are zero when $l' \ne l$ or $m' \ne m$, and that the nonzero coefficients are those described in the Lemma as $\Gamma_{l,m,2}$.

@@

@@

			\big(\vec{\eta}_{2}\big)_{j-1} & j = 2(2l+1), 6(2l+1) - 4 \\

>>>>

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. It is useful to be able to transform between these cartesian coordinates and the spherical coordinates ($\varphi, \theta$). Throughout this document we will use the convention that the spherical coordinate angles be defined by
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi.
}
where $\rho(z) := \sqrt{1-z^2}$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}:
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}.
}
The spherical harmonics as defined are then orthonormal with respect to the inner product
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in $\R^3$.

@@

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jx}
		\centering
		\caption{$J_x$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jy}
		\centering
		\caption{$J_y$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\includegraphics[scale=0.5]{spy-of-Jz}
		\centering
		\caption{$J_z$}
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of the Jacobi operator matrices, showing their sparsity and structure, up to order $N=10$.}
	\label{fig:sh-spy-jacobi}
\end{figure}

Using \bsrefeqns{eqn:shrecx}{eqn:shrecz}, we have that the Jacobi matrices take the following block-tridiagonal form (i.e. have block-bandwidths $(1,1)$):
\bseqn{
	J_x &= 
		\begin{pmatrix}
			B_{x,0} & A_{x,0} & & & & \\
			C_{x,1} & B_{x,1} & A_{x,1} & & & \\
			& C_{x,2} & B_{x,2} & A_{x,2}  & & & \\
			& & C_{x,3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}, \\
	J_y &= 
		\begin{pmatrix}
			B_{y,0} & A_{y,0} & & & & \\
			C_{y,1} & B_{y,1} & A_{y,1} & & & \\
			& C_{y,2} & B_{y,2} & A_{y,2}  & & & \\
			& & C_{y,3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}, \\
	J_z &= 
		\begin{pmatrix}
			B_{z,0} & A_{z,0} & & & & \\
			C_{z,1} & B_{z,1} & A_{z,1} & & & \\
			& C_{z,2} & B_{z,2} & A_{z,2}  & & & \\
			& & C_{z,3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}.
}
$J_x$ has subblock-bandwidths $(2,2)$, where the blocks for $l \in \No$ are given by:
\bseqn{
	A_{x,l} &:= 
		\begin{pmatrix}
			\alpha_{l,-l,3} & 0 & \alpha_{l,-l,4} & & \\
			& \ddots & \ddots & \ddots & \\
			& & \alpha_{l,l,3} & 0 & \alpha_{l,l,4} \\
		\end{pmatrix} \in \R^{(2l+1)\times(2l+3)}, \\
	B_{x,l} &:= 0 \in \R^{(2l+1)\times(2l+1)} \\
	C_{x,l} &:= 
		\begin{pmatrix}
			\alpha_{l,-l,2} & & \\
			0 & \ddots & \\
			\alpha_{l,-l+2,1} & \ddots & \alpha_{l,l-2,2} \\
			& \ddots & 0 \\
			& & \alpha_{l,l,1}
		\end{pmatrix} \in \R^{(2l+1)\times(2l-1)} \quad (l \ne 0).
}
$J_y$ also has subblock-bandwidths $(2,2)$, where the blocks for $l \in \No$ are given by:
\bseqn{
	A_{y,l} &:= 
		\begin{pmatrix}
			\beta_{l,-l,3} & 0 & \beta_{l,-l,4} & & \\
			& \ddots & \ddots & \ddots & \\
			& & \beta_{l,l,3} & 0 & \beta_{l,l,4} \\
		\end{pmatrix} \in \C^{(2l+1)\times(2l+3)}, \\
	B_{y,l} &:= 0 \in \R^{(2l+1)\times(2l+1)} \\
	C_{y,l} &:= 
		\begin{pmatrix}
			\beta_{l,-l,2} & & \\
			0 & \ddots & \\
			\beta_{l,-l+2,1} & \ddots & \beta_{l,l-2,2} \\
			& \ddots & 0 \\
			& & \beta_{l,l,1}
		\end{pmatrix} \in \C^{(2l+1)\times(2l-1)} \quad (l \ne 0).
}
Finally, $J_z$ has subblock-bandwidths $(1,1)$, where the blocks for $l \in \No$ are given by:
\bseqn{
	A_{z,l} &:= 
		\begin{pmatrix}
			0 & \gamma_{l,-l,2} & 0 & & \\
			& \ddots & \ddots & \ddots & \\
			& & 0 &  \gamma_{l,l,2} & 0 \\
		\end{pmatrix} \in \R^{(2l+1)\times(2l+3)}, \\
	B_{z,l} &:= 0 \in \R^{(2l+1)\times(2l+1)} \\
	C_{z,l} &:= 
		\begin{pmatrix}
			0 & & \\
			\gamma_{l,-l+1,1} & \ddots & \\
			0 & \ddots & 0 \\
			& \ddots & \gamma_{l,l-1,1} \\
			& & 0
		\end{pmatrix} \in \R^{(2l+1)\times(2l-1)} \quad (l \ne 0).
}
The sparsity and structure of the Jacobi matrices as we have defined them can be seen visually in \bsreffig{fig:sh-spy-jacobi}.

@@

	A_{l,m,5} &:= {i \over 2l(l+1)}

@@

			i \: d_{l,\abs{m}} \: {\clm \over 2 c_l^{m-1}} &\text{if } m \le 0

@@

	A_{l,m,6} &:= {i \over 2l(l+1)}

@@

			- d_{l,m} \: {\clm \over c_l^{m+1}} &\text{if } m \ge 0 \\
			d_{l,\abs{m} - 1} \: {c_l^{m+1} \over \clm} &\text{if } m < 0

@@

	B_{l,m,j} &:= i (-1)^{j+1} A_{l,m,j} \quad \text{for } j = 1,\dots,6 \\

@@

	\Gamma_{l,m,3} &:= {i \: m \over l(l+1)}

@@

	Fix $l \in \N$ and $m \in \Z$ s.t. $-l \le m \le l$. The proof we will use is to directly calculate the nonzero coefficients in each expansion using inner products. To this end, note that
\bseqn{
	x \gradYlm = \sum_{l'=1}^{l+1} \sum_{m'=-l'}^{l'} \calA_{l',m'} \gradYlmprime + \calA^\perp_{l',m'} \gradpYlmprime
}
where the coefficients are given by
\bseqn{
	\calA_{l',m'} = {\ip<x \gradYlm, \gradYlmprime> \over \norm{\gradYlmprime}^2} = {1 \over l'(l'+1)} \ip<x \gradYlm, \gradYlmprime> \\
	\calA^\perp_{l',m'} = {\ip<x \gradYlm, \gradpYlmprime> \over \norm{\gradpYlmprime}^2} = {1 \over l'(l'+1)} \ip<x \gradYlm, \gradpYlmprime>.
}
The aim will be to show that the only nonzero coefficients $\calA_{l',m'}$, $\calA^\perp_{l',m'}$ match the $A_{l,m,j}$ for $j=1,\dots,6$ in the Lemma. Now, using a change of variable, we have that
\bseqnnumber{
	&\ip<x \gradYlm, \gradYlmprime> \nonumber \\
	&= \iint_\Omega \costheta \sinphi \: \gradYlm \cdot {\gradYlmprime}^* \sinphi \: \D\phi \: \D\theta \nonumber \\
	&= \clm \clmprime \Big( \int_0^{2pi} \eimtheta \eimthetaprime \costheta \: \D\theta \Big) \nonumber \\
	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ \big[ \abs{m} z \jaclmabs{l}{m}(z) - \rho(z)^2 \jaclmabsdz{l}{m}(z) \big] \big[ \abs{m'} z \jaclmabs{l'}{m'}(z) - \rho(z)^2 \jaclmabs{l'}{m'}(z) \big] \nonumber \\
	&\quadeight + mm' \jaclmabs{l}{m}(z) \jaclmabs{l'}{m'}(z) \Big\} \: \rho(z)^{\abs{m}+\abs{m'}-1} \: \D z \Big) \nonumber \\
	&= \clm \clmprime \pi (\delta_{m',m-1} + \delta_{m',m+1}) \nonumber \\
	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ \jaclmabs{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m}+\abs{m'}-1} \big(mm' + \abs{m}\abs{m'} z^2 \big) \nonumber \\
	&\quadeight - \abs{m} z \jaclmabs{l}{m} \jaclmabsdz{l'}{m'} \rho^{\abs{m}+\abs{m'}+1} \nonumber \\
	&\quadeight - \abs{m'} z \jaclmabsdz{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m}+\abs{m'}+1} \nonumber  \\
	&\quadeight + \jaclmabsdz{l}{m} \jaclmabsdz{l'}{m'} \rho^{\abs{m}+\abs{m'}+3} \Big\} \: \D z \Big). \label{eqn:lemmaVSHrecurrencesproof}
}
We can conclude that the only possible nonzero coefficients are when $m' = m \pm 1$, and so it is left to evaluate the integral that is left at \bsrefeqn{eqn:lemmaVSHrecurrencesproof} when $m'$ takes either value. Since we need to take account of the fact that $m$ can be negative, to simplify our argument we make use of the fact that $m' = m - 1$ for $m < 0$ is analogous to the case $m' = m+1$ for $m > 0$, and vice versa.

@@

	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ \sgn{m} 2 i \, m \, m' \, z \jaclmabs{l}{m} \jaclmabs{l'}{m'} \rho^{\abs{m} + \abs{m'} - 1} \nonumber \\

@@

	&\quadeight - i m' \jaclmabs{l}{m} \jaclmabsdz{l'}{m'} \rho^{\abs{m} + \abs{m'} +1} \Big\} \: \D z \Big), \label{eqn:lemmaVSHrecurrencesproof4}

@@

since when the above is nonzero, $m, m'$ will have the same sign (or one is zero) and so $\abs{m}m' + \abs{m'}m = \sgn{m}mm'$. Once again, we can conclude that $\calA^\perp_{l',m'}$ are zero for $m' \notin \{m-1, m+1\}$. It is left to evaluate the integral that is left at \bsrefeqn{eqn:lemmaVSHrecurrencesproof4} when $m'$ takes either $m\pm 1$ value. First, consider the case that $m > 0$ and $m' = m - 1$. The integral is then, using integration by parts:

@@

@@

	&= - \int_{-1}^{1} i \; d_{l',m-1} \; \jaclm{l}{m} \jaclm{l'}{m} \rho^{2m} \: \D z \nonumber \\

@@

@@

	&= \int_{-1}^{1} i \jaclmabs{l}{m} \jaclmbracketdz{l'}{\abs{m} - 1} \rho^{2\abs{m}} \: \D z \nonumber \\
	&= \int_{-1}^{1} i \; d_{l',\abs{m}-1} \; \jaclmabs{l}{m} \jaclmabs{l'}{m} \rho^{2\abs{m}} \: \D z \nonumber \\
	&= i \; {d_{l,\abs{m}-1} \over 2 \pi (c_l^{\abs{m}})^2} \; \delta_{l',l}. \label{eqn:lemmaVSHrecurrencesproof6}

@@

where the coefficients are given by

@@

The integral over $\theta$ in \bsrefeqn{eqn:lemmaVSHrecurrencesproofz1} is simply $2 \pi \delta_{m',m}$ showing the above is zero when $m' \ne m$, and thus so will the coefficients $\calC_{l',m'}$ be when this is the case. Focusing on the second integral (over $z$) when $m' = m$, this becomes:

@@

We can see here in \bsrefeqn{eqn:lemmaVSHrecurrencesproofz3} that the nonzero coefficients $\calC_{l',m'}$ are those described in the Lemma as $\Gamma_{l,m,1}$ and $\Gamma_{l,m,2}$.

Next, we also have that
\bseqnnumber{
	&\ip<z \gradYlm, \gradpYlmprime> \nonumber \\
	&= \iint_\Omega \cosphi \: \gradYlm \cdot {\gradpYlmprime}^* \sinphi \: \D\varphi \: \D\theta \nonumber \\
	&= \clm \clmprime \Big( \int_0^{2pi} \eimtheta \eimthetaprime \: \D\theta \Big) \nonumber \\
	&\quadfour \cdot \Big( \int_{-1}^{1} \Big\{ i m' z \jaclmabs{l'}{m'}(z) \rho(z)^{\abs{m} + \abs{m'} - 2} \big[ \abs{m} z \jaclmabs{l}{m}(z) - \rho(z)^2 \jaclmabsdz{l}{m}(z) \big] \nonumber \\
	&\quadeight + i m z \jaclmabs{l}{m}(z) \rho(z)^{\abs{m} + \abs{m'} - 2} \big[ \abs{m'} z \jaclmabs{l'}{m'}(z) - \rho(z)^2 \jaclmabs{l'}{m'}(z) \big] \Big\} \: \D z \Big) \nonumber \\
	&= \clm c_{l'}^{m} 2 \pi \delta_{m',m} \: \int_{-1}^{1} \Big\{ 2i m\abs{m} z^2 \jaclmabs{l}{m}(z) \jaclmabs{l'}{m}(z) \rho(z)^{2\abs{m} - 2} \nonumber \\
	&\quadeight \quadtwo - i m z \jaclmabs{l}{m}(z) \jaclmabsdz{l'}{m}(z) \rho(z)^{2\abs{m}} \nonumber \\
	&\quadeight \quadtwo - i m z \jaclmabsdz{l}{m}(z) \jaclmabs{l'}{m}(z) \rho(z)^{2\abs{m}} \Big\} \: \D z \nonumber \\
	&= \clm c_{l'}^{m} 2 \pi \delta_{m',m} \: \int_{-1}^{1} i m \jaclmabs{l}{m}(z) \jaclmabs{l'}{m}(z) \rho(z)^{2\abs{m}} \: \D z \nonumber \\
	&= \delta_{l',l} \delta_{m',m} i m. \label{eqn:lemmaVSHrecurrencesproofz4}
}
%&= \delta_{l',l} \delta_{m',m} i m (-1)^m. \label{eqn:lemmaVSHrecurrencesproofz4}
Again, we can see here in \bsrefeqn{eqn:lemmaVSHrecurrencesproofz4} that the coefficients $\calC^\perp_{l',m'}$ are zero when $l' \ne l$ or $m' \ne m$, and that the nonzero coefficients are those described in the Lemma as $\Gamma_{l,m,3}$.

@@

The sparsity and structure of the Jacobi matrices for the VSHs as we have defined them can be seen visually in \bsreffig{fig:vsh-spy-jacobi}.

@@

			\big(\vec{\eta}_{2}\big)_{j-1} & j = 2(2l+1), 6(2l+1) - 2 \\
